---
title: "Polar bear example: random slopes and hierarchical smooths"
author: "Natasha Klappstein"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r options, include = FALSE}
knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(
  message = FALSE, error = FALSE, warning = FALSE,
  comment = NA
)
```

## Set-up
First, we will load the packages we need for the analysis. `mgcv` is the main package for fitting the SSFs. `gratia` is full of useful tools for visualising and extracting `mgcv` outputs.
```{r packages}
library(mgcv)
library(cowplot)
library(ggplot2)
library(gratia)
```

We will be analysing GPS locations from 13 adult polar bears in the Beaufort Sea (provided by Andrew Derocher). The data have already been processed to include random steps, and the `stratum` column indicates the step ID and `obs` indicates whether the location is an observed location (1) or random location (0). `step` are the step lengths (km), `angle` are turning angles (radians), and `ice_conc` is the ice concentration (\%).

```{r pb_data}
# load location data
data <- readRDS("data/polar_bear.RData")
head(data)
```

The objective of the analysis is to account for inter-individual variability in selection for ice concentration using two methods: i) a random slope model which assumes linear patterns of selection for all individuals, and ii) a hierarchical smooth which allows individuals to have non-linear patterns of selection. For both, we need to define a variable for times, which can be set to 1 for all rows.
```{r}
# create a dummy variable for times
data$times <- 1
```

## Fit random slope model

First, we will fit a random slope model. We are modelling step lengths with a gamma distribution (so we include step and its log as covariates), and turning angles with a von Mises distribution (so we include the cosine of the turning angle). Then, we include a global (linear) term for ice concentration, as well as the random slopes (with `s(ice_conc, ID, bs = "re")`). 
```{r slopes, cache=TRUE}

# fit with random slopes
fit_slopes <- gam(cbind(times, stratum) ~ 
                    step + 
                    log(step) + 
                    cos(angle) +
                    ice_conc + 
                    s(ice_conc, ID, bs = "re") , 
                  data = data,
                  family = cox.ph, 
                  weights = obs)
```

We can look at the summary of the model object.
```{r summary_slopes}
summary(fit_slopes)
```

We may also want to visualise the relationship, as well as the estimated slopes for each individual. The basic plotting functions in `gratia` and `mgcv` can easily plot the population-level slope, but it requires custom code to plot each random slope. Below, I predict for each individual and plot. I've hidden the actual plotting code in `ggplot2`, as it's quite long but you can check out the code file for that if you need. 

```{r plot_slopes}
# define grid of ice concentration to predict over
conc_grid <- seq(0, 100, 1)

# get slopes for each individual
bears <- unique(data$ID)
coefID <- as.vector(coef(fit_slopes)[5:(4+length(bears))]) + coef(fit_slopes)[4]

# predict for each individual
r_slope_ID <- NULL
for(i in 1:13) {
  r_slope <- data.frame(bearID = bears[i], conc = conc_grid, RSS = exp(coefID[i] * conc_grid))
  r_slope_ID <- rbind(r_slope_ID, r_slope)
}

```

```{r, fig.height=3.5, fig.width=8, echo = FALSE}
# create data frame for plotting
pop_slope <- data.frame(conc = conc_grid, 
                        RSS = exp(coef(fit_slopes)[4] * conc_grid), 
                        lower = exp((coef(fit_slopes)[4] - 0.001515 * 1.96) *conc_grid),
                        upper = exp((coef(fit_slopes)[4] + 0.001515 * 1.96) *conc_grid))

# population estimate
p1 <- ggplot(data = pop_slope, aes(x = conc, y = RSS)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey70", alpha = 0.5) +
  geom_line(color = "black", size = 0.75) + 
  ggtitle("") +
  ylim(c(0,6)) +
  ylab(expression(exp(beta[ice] * ice))) +
  xlab("Ice concentration (%)") +
  theme_bw() + 
  theme(legend.position = "none")

# individual-level estimates
p2 <- ggplot(data = pop_slope, aes(x = conc, y = RSS)) + 
  geom_line(data = r_slope_ID, color = "firebrick", alpha = 0.5,
            aes(x = conc, y = RSS, group = bearID))+
  ggtitle("") +
  ylim(c(0,6)) +
  ylab(expression(exp(beta[ice]))) +
  xlab("Ice concentration (%)") +
  theme_bw() + 
  theme(legend.position = "none")
plot_grid(p1, p2)
```



```{r smooths, cache=TRUE}
# hierarchical smooths
fit_smooths <- gam(cbind(times, stratum) ~ 
                     step + 
                     log(step) + 
                     cos(angle) +
                     s(ice_conc, k = 5) + 
                     s(ice_conc, ID, k = 5, bs = "fs"),
                   data = data,
                   method = "REML",
                   family = cox.ph, 
                   weights = obs)
```

```{r summary_sm}
summary(fit_smooths)
```

```{r}
###############################
# make plot of random smooths #
###############################
r_smooths <- smooth_estimates(fit_smooths)
pop_smooth <- r_smooths[c(1:100),]

r_smooth_ID <- NULL
for(i in 1:length(bears)) {
  
  smooth_sub <- subset(r_smooths, 
                       smooth == "s(ice_conc,ID)" & ID == bears[i])
  
  smooth_sub <- data.frame(ID = bears[i], 
                           ice_conc = smooth_sub$ice_conc,
                           est = smooth_sub$est + pop_smooth$est)
  
  r_smooth_ID <- rbind(r_smooth_ID, smooth_sub)
}
```


```{r, fig.height=4.5, fig.width=10, echo = FALSE}
p3 <- ggplot(data = pop_smooth, aes(x = ice_conc, y = exp(est))) + 
  geom_ribbon(aes(ymin = exp(est - 1.96 * se), ymax = exp(est + 1.96 * se)), 
              fill = "grey70", alpha = 0.5) +
  geom_line(color = "black", size = 0.75) + 
  ylim(c(0, 2.35)) +
  ggtitle("") +
  ylab(expression(exp(f(ice)))) +
  xlab("Ice concentration (%)") +
  theme_bw() + 
  theme(legend.position = "none")

p4 <- ggplot(data = pop_smooth, aes(x = ice_conc, y = exp(est))) + 
  geom_line(data = r_smooth_ID, color = "royalblue", alpha = 0.6,
            aes(x = ice_conc, y = exp(est), group = ID))+
  ylim(c(0, 2.35)) +
  ggtitle("") +
  ylab(expression(exp(f(ice)))) +
  xlab("Ice concentration (%)") +
  theme_bw() + 
  theme(legend.position = "none")
plot_grid(p3, p4)
```