---
title: "Simulated data: spatial smoothing"
author: "Natasha Klappstein"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r options, include = FALSE}
knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(
  message = FALSE, error = FALSE, warning = FALSE,
  comment = NA
)
```

This document contains code to reproduce the spatial smoothing example in Appendix B of Klappstein et al. (2024). Note that these results may be slightly different to those in the paper, as we used the `hmmSSF` package in the original analysis, but are using `amt` here for reader clarity. The general idea is to use a spatial smooth to capture select for an ``unknown" centre of attraction.

## Data and setup

```{r packages}
# load packages

library(mgcv)
library(gratia)
library(ggplot2)
library(wesanderson)
library(dplyr)
library(terra)
library(cowplot)
library(amt)
```

The data were simulated from a step selection function with selection for two covariates: i) one randomly generated covariate (called `cov`) with selection $\beta_{cov} = 5$, and ii) distance to a centre of attraction at coordinates (10, 10) (denoted `centre`) with selection $\beta_{centre} = -0.075$. We can load both of these data sources (location and covariate) and translate the covariates back into raster format.

```{r data}
# load and view location data
data <- readRDS("data/sim_data.RData")
head(data)

# load and process raster data
cov_data <- readRDS("data/cov_data.RData")
cov <- rast(cov_data[[1]], type = "xyz")
centre <- rast(cov_data[[2]], type = "xyz")
names(cov) <- "cov"
```

Below are plots of the simulated locations overlaid onto each covariate raster.

```{r plot_raster, echo = FALSE}
library(tidyterra)
p1 <- ggplot() + 
  geom_spatraster(data = cov, aes(fill = cov)) +
  coord_equal() +
  scale_fill_viridis_c(option = "mako", name = "covariate") + 
  ylim(c(min(data$y) -4, max(data$y)+4)) +
  xlim(c(min(data$x)-4, max(data$x)+4)) + 
  xlab("x") + ylab("y") + 
  geom_point(data = data, aes(x = x, y = y), 
             alpha = 0.2, size = 0.1)

p2 <- ggplot() + 
  geom_spatraster(data = centre, aes(fill = focal_sum)) +
  coord_equal() +
  scale_fill_viridis_c(option = "mako", limits = c(0, 41), name = "distance") + 
  ylim(c(min(data$y) -4, max(data$y)+4)) +
  xlim(c(min(data$x)-4, max(data$x)+4)) +
  xlab("x") + ylab("y") +
  geom_point(data = data, aes(x = x, y = y), 
             alpha = 0.2, size = 0.1)

plot_grid(p1, p2)

```

## Generating random points

In this example, we will use the `amt` package to generate random points. First, we need to transform our dataset into class `track_xy*`, then we can use `random_steps` to generate 20 random points for each observed location. We use the amt default, which uses a gamma distribution for random step lengths.

```{r}

# create track_xy*
track <- make_track(data, x, y) |> 
  steps()

# get random steps
set.seed(55)
ssf_data <- random_steps(track, n = 20)

# get covariate values
ssf_data <- extract_covariates(ssf_data, cov)
head(ssf_data)

```

## Model fitting

Then, we can add a constant `times` variable for the Cox PH implementation and fit the model in `mgcv`. Our response is `times` in conjunction with the stratum ID (`step_id_`). We're including linear effects for step length `sl_` and the habitat covariate `cov`. However, rather than including the centre of attraction, we're assuming that we do not have that information. Instead, we will include a spatial smooth to account for any remaining spatial heterogeneity not explained by `cov`.

```{r fit, cache = TRUE}
# add dummy variable for times
ssf_data$times <- 1

# fit in mgcv
fit <- gam(cbind(times, step_id_) ~ 
              sl_ + 
              cov +
              s(x2_, y2_), 
            data = ssf_data,
            family = cox.ph, 
            weights = case_)
```

```{r}
summary(fit)
```

## Model interpretation

Our summary indicates that we estimate the effect of the habitat covariate quite well, as the estimate is close to the truth ($\beta_{cov} = 5$). We can also plot the spatial smooth quickly with `gratia`.
```{r plot}
gratia::draw(fit, rug = FALSE)
```

We can also reproduce the plot in the paper with the following code, which allows us to better visualise how the spatial smooth is capturing the centre of attraction.

```{r plot_model, fig.dim = c(9, 4.5)}

# get spatial smooth estimates 
spatial <- smooth_estimates(fit, smooth = "s(x2_,y2_)")

# plot covariate
p1a <- ggplot() + 
  geom_spatraster(data = cov, aes(fill = cov)) +
  coord_equal() +
  scale_fill_viridis_c(option = "mako", name = "covariate") + 
  ylim(c(min(data$y) -4, max(data$y)+4)) +
  xlim(c(min(data$x)-4, max(data$x)+4)) +
  xlab("x") + ylab("y") +
  xlab("x") + ylab("y")

# plot centre of attraction
p2a <- ggplot() + 
  geom_spatraster(data = centre, aes(fill = focal_sum)) +
  coord_equal() +
  scale_fill_viridis_c(option = "mako", limits = c(0, 41), name = "distance") + 
  ylim(c(min(data$y) -4, max(data$y)+4)) +
  xlim(c(min(data$x)-4, max(data$x)+4)) +
  xlab("x") + ylab("y") +
  xlab("x") + ylab("y")

# plot spatial smooth
p3 <- ggplot(spatial, aes(x = x2_, y = y2_, fill = est)) + 
  geom_raster() + coord_equal() +
  scale_fill_viridis_c(option = "rocket") + 
  geom_point(data = data, aes(x = x, y = y, fill = ID), 
             alpha = 0.2, size = 0.1)

# plot altogether
plot_grid(p1a, p2a, p3, nrow = 1)
```







